Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-01 doc/paper-lcr.tex
Conclusions in bullet points, commenting colors and styles fixed, with todo macro added

Previous attempts at reconstructing the line of sight mass distribution in
strong lens fields have focused on spectroscopic surveys to ascertain group
membership, and on understanding the external shear apparently required by the
strong lens models.  The probability density function (PDF) $\Pr(\xkappa)$ is
the target for any reconstruction analysis, and the key quantity for any
ensemble analysis, since by marginalising over it we can correctly propagate
the uncertainties due to weak lensing by line of sight mass structure. 
\flag{More needed here. Previous work on environments: Keeton, Zabludoff,Auger, Momcheva, Wong, others.}
Most recently, \citet{Wong++2011} ran Monte
Carlo reconstructions of the mass in nine lens fields, and compared the
resulting predicted shear distributions with the lens model external shear.
Their simulations show most of  the shear being generated by bright galaxies
within 2~arcminutes of the lens, and that both line of sight structures and
the group of galaxies in each lens plane contribute significant proportions of
the shear. The significant discrepancies between the lens model shear and the
Monte Carlo predictions are attributed to both the environment modelling and
the strong lens modelling itself.

These observational investigations have provided very useful insight into the
problem.  Coming from the opposite direction, large numerical simulations have
been used to predict, from first principles the distribution of likely
external shears and convergences along strong lens lines of sight. \flag{More
needed here. Holder \& Schechter, Dalal \& Watson}. \citet{Hilbert++2009}
performed ray-tracing experiments through the Millenium Simulation, generating
a predicted $\xkappa$ at every position in a simulated sky. The first step
towards using this excellent calibration resource was taken by
\citet{Suyu++2010}, who selected Millenium lines of sight that a) were found
to contain a strong lens and b) had comparable apparent galaxy overdensity in
a 45 arcsec radius aperture down to $i < 24$. This overdensity could be
compared to the overdensity of the B1608$+$656 line of sight
\citep{Fass+Koop2010}, and hence an estimate of  $\Pr(\xkappa)$ could be
made.

@@

If the intrinsic source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence (and appropriate re-scaling of the lens potential) the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in the lens modelling, if cosmological parameters are to be estimated accurately and
precisely from observed time delays \citep{suyu2010}.

@@

\flag{The all-important calibration of the reconstruction has to be done
against Stefan's ray-traced kappas though -- so it is these that we have to
predict PDFs for, first and foremost. THEN we have to consider the question of
which kappa to use to correct the time delay distances: Stefan's kappa (as in
B1608 or RXJ1131), or kappa-keeton.}

@@

Cosmological dark matter simulations can be used to generate mock lines of
sight...
\flag{Stefan's section on how the ray tracing was done and how halos are extracted from
 the MS n-body particles}

@@

Unphysically, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line of sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}. Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_{\rm NFW} (r)}{1+\left(r/r_{t}\right)^2},
\ee
which is the same as the NFW profile in the limit that $r_t$ goes to infinity. The convergence and shear for the BMO profile are given in Appendix \ref{appendix:halos}.

@@

We choose 100000 lines of sight from random positions in the catalogue, and for each line of sight we reconstruct the
convergences from every halo within 5 arcminutes and compare the reconstructed
$\kapparec$ with those from the ray tracing results. The standard deviation on $\kapparay$-$\kapparec$ 
shown in Figure \ref{fig:ScattervsTruncation} as a function of the radius at which the halo is truncated.
 It is clear from Figure \ref{fig:ScattervsTruncation} 
that as long as the truncation is greater than a few virial radii, the choice of truncation
radius does not broadly affect the scatter on $\kapparay-\kapparec$; from now on we choose 
to truncate the halos at $5 \Rhalo$. Since $\sigma\left(\kapparay-\kapparec\right)$ does not go to
zero, it is clear that out halo based prescription does not perfectly replicate the n-body results, we refer to this
 as the intrinsic error of the prescription.
 The intrinsic error has a standard deviation of $\sigi \sim 0.018$, although
this figure is inflated by a few outliers in the tails of $\pr\left(\kapparay-\kapparec\right)$. Such outliers would 
likely be rejected by their reconstructed shear not matching the shear of the lens model 
(see \citet{wong} for the practicalities of this);
 this makes our estimate of the intrinsic error pessimistic. We do not include any additional constraints on P($\kappax$), given prior knowledge of the external shear in this work.

@@

The method we have outlined can also be used to estimate the external shear along a line of sight. Shear is an observable that can be extracted from strong lens modelling, however there is a degeneracy between internal and external shear. Progress has been made in disentangelling external and internal shear \citep[\eg][]{xxx} but there are still significant uncertainties: \citet{Wong} attempted to match the shear from strong lens models with a reconstruction of the local lens group environment, but found a tension between the strong lens model and the reconstruction of the environment. Since \citet{Wong} only reconstructed the local lens group, rather than the full line of sight contribution, it is unclear whether the external shear from lens models can be reconciled with a line of sight reconstruction. {\it If} external shear can be measured, it provides an additional constraint on which of the Millenium Simulation lines of sight are similar to the reconstructed line of sight. \citet{suyu2012b} found that for RXJ1131 combining shear constriants with galaxy number count overdensity, gave a significantly different $\pr(\kappax|\gamma,N_{45})$ copmpared to the PDFfrom number count overdensity alone, $\pr(\kappax|N_{45})$.

@@

In this work we have taken mock lines of sight from the Millenium Simulation and estimated the external convergence along those lines of sight using a simple halo model prescription. We have used the ray traced convergence as a calibration, allowing us to relate the total convergence along a line of sight with the sum of convergences from each halo in our catalogue. Our model uses a truncated spherical NFW profile for each dark matter halo and neglects voids, but despite the model's simplicity the reconstructed $\sumkappah$ is a good tracer of $\kappax$. Using Millenium Simulation lightcones allowed us to test the reconstruction process given perfect knowledge of halo mass and redshift. We found that with perfect knowledge of the halo mass and redshift, the reconstruction gives an unbiased estimator of $\pr(\kappax)$ that is almost almost a factor of 2 less broad to the global $\pr(\kappax)$. For the most overdense lines of sight, the reconstruction produces a very broad PDF, but since the reconstruction can be performed before follow-up time is invested it is possible to discard the most uncertain lines of sight and prevent the waste of follow-up time.

Having found that the reconstruction process is effective given perfect knowledge of halo mass and redshift, we investigated the effects of reasonable uncertainties in the stellar mass and redshift of each halo, and propogated these uncertainties into a $\pr\left(\sumkappah\right)$ for each line of sight. With complete spectroscopic redshift coverage we find that the median of $\pr\left(\sumkappah\right)$ can be converted into an estimate of $\pr(\kappax)$; the PDFS are roughly 10\% broader than given perfect knowledge of halo mass and redshift; given only photometric redshifts (which in turn causes much less certain stellar masss estimates) causes another $\sim$10\% increase to the width of $\pr(\kappax)$.

We found that it is very rare for halos further than 2 arcminutes to make a significant contribution to $\kappax$. We also found that including halos whose host galaxy is less luminous than $i=24$ does not significantly improve our reconstruction proceedure. As such a photometric survey of a 4$\times$4 arcminute patch around the lens gets close to the limiting uncertainties of our simple reconstruction recipe. Such a reconstruction produces $\pr(\kappax)$ that have a meanwidth of 0.0163 which is a 1.5 times less broad than the global $\pr(\kappax)$, but by following-up only the most certain cosmological value per lens can be increased further. We find that the lines of sight with the sharpest $\pr(\kappax)$ are typically under-dense.

With a photometric reconstruction of all lens fields, and following up only the lenses with the most constraining $\pr(\kappax)$, the width of the dominant statistical uncertainty in time-delay cosmography can be halved, whilst at the same time decreaing any potentail for a systematic error due to lenses being biased in $\kappax$.

@@

\flag{SHEAR HERE}

>>>>

Previous attempts at reconstructing the line of sight mass distribution in
strong lens fields have focused on spectroscopic surveys to ascertain group
membership, and on understanding the external shear apparently required by the
strong lens models.  The probability density function (PDF) $\Pr(\xkappa)$ is
the target for any reconstruction analysis, and the key quantity for any
ensemble analysis, since by marginalising over it we can correctly propagate
the uncertainties due to weak lensing by line of sight mass structure. 
\todo{Tom, Matt}{More needed here. Previous work on environments: Keeton,
Zabludoff,Auger, Momcheva, Wong, others.}
Most recently, \citet{Wong++2011} ran Monte
Carlo reconstructions of the mass in nine lens fields, and compared the
resulting predicted shear distributions with the lens model external shear.
Their simulations show most of  the shear being generated by bright galaxies
within 2~arcminutes of the lens, and that both line of sight structures and
the group of galaxies in each lens plane contribute significant proportions of
the shear. The significant discrepancies between the lens model shear and the
Monte Carlo predictions are attributed to both the environment modelling and
the strong lens modelling itself.

These observational investigations have provided very useful insight into the
problem.  Coming from the opposite direction, large numerical simulations have
been used to predict, from first principles the distribution of likely
external shears and convergences along strong lens lines of sight. \todo{Phil}{More
needed here. Holder \& Schechter, Dalal \& Watson}. \citet{Hilbert++2009}
performed ray-tracing experiments through the Millenium Simulation, generating
a predicted $\xkappa$ at every position in a simulated sky. The first step
towards using this excellent calibration resource was taken by
\citet{Suyu++2010}, who selected Millenium lines of sight that a) were found
to contain a strong lens and b) had comparable apparent galaxy overdensity in
a 45 arcsec radius aperture down to $i < 24$. This overdensity could be
compared to the overdensity of the B1608$+$656 line of sight
\citep{Fass+Koop2010}, and hence an estimate of  $\Pr(\xkappa)$ could be
made.

@@

If the intrinsic source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence (and appropriate re-scaling of the lens potential) the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in the lens modelling, if cosmological parameters are to be estimated accurately and
precisely from observed time delays \citep{Suyu2010}.

@@

\tom{The all-important calibration of the reconstruction has to be done
against Stefan's ray-traced kappas though -- so it is these that we have to
predict PDFs for, first and foremost. THEN we have to consider the question of
which kappa to use to correct the time delay distances: Stefan's kappa (as in
B1608 or RXJ1131), or kappa-keeton.}

@@

Cosmological dark matter simulations can be used to generate mock lines of
sight...
\todo{Stefan}{Write section on how the ray tracing was done and how halos are extracted from
 the MS n-body particles}

@@

Unphysically, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line of sight will also be divergant\footnote{At large radius, $\Sigma_{\rm
nfw} \propto R^{-2}$, but the differential number of halos centred within an
annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$,
so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which
diverges logarithmically}. Since infinite mass is unphysical, the profile must
be truncated at some point. \todo{Tom}{Add notes on previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_{\rm NFW} (r)}{1+\left(r/r_{t}\right)^2},
\ee
which is the same as the NFW profile in the limit that $r_t$ goes to infinity. The convergence and shear for the BMO profile are given in Appendix \ref{appendix:halos}.

@@

We choose 100000 lines of sight from random positions in the catalogue, and for each line of sight we reconstruct the
convergences from every halo within 5 arcminutes and compare the reconstructed
$\kapparec$ with those from the ray tracing results. The standard deviation on $\kapparay$-$\kapparec$ 
shown in Figure \ref{fig:ScattervsTruncation} as a function of the radius at which the halo is truncated.
 It is clear from Figure \ref{fig:ScattervsTruncation} 
that as long as the truncation is greater than a few virial radii, the choice of truncation
radius does not broadly affect the scatter on $\kapparay-\kapparec$; from now on we choose 
to truncate the halos at $5 \Rhalo$. Since $\sigma\left(\kapparay-\kapparec\right)$ does not go to
zero, it is clear that out halo based prescription does not perfectly replicate the n-body results, we refer to this
 as the intrinsic error of the prescription.
 The intrinsic error has a standard deviation of $\sigi \sim 0.018$, although
this figure is inflated by a few outliers in the tails of $\pr\left(\kapparay-\kapparec\right)$. Such outliers would 
likely be rejected by their reconstructed shear not matching the shear of the lens model 
(see \citet{Wong} for the practicalities of this);
 this makes our estimate of the intrinsic error pessimistic. We do not include any additional constraints on P($\kappax$), given prior knowledge of the external shear in this work.

@@

The method we have outlined can also be used to estimate the external shear
along a line of sight. Shear is an observable that can be extracted from
strong lens modelling, however there is a degeneracy between internal and
external shear. Progress has been made in disentangelling external and
internal shear \citep[\eg][]{xxx} but there are still significant
uncertainties: \citet{Wong} attempted to match the shear from strong lens
models with a reconstruction of the local lens group environment, but found a
tension between the strong lens model and the reconstruction of the
environment. Since \citet{Wong} only reconstructed the local lens group,
rather than the full line of sight contribution, it is unclear whether the
external shear from lens models can be reconciled with a line of sight
reconstruction. {\it If} external shear can be measured, it provides an
additional constraint on which of the Millenium Simulation lines of sight are
similar to the reconstructed line of sight. \citet{Suyu2012b} found that for RXJ1131 combining shear constriants with galaxy number count overdensity, gave a significantly different $\pr(\kappax|\gamma,N_{45})$ copmpared to the PDFfrom number count overdensity alone, $\pr(\kappax|N_{45})$.

@@

In this work we have investigated a simple halo model prescription for
reconstructing all the mass along a line of sight to an intermediate redshift
source. We have used the ray-traced lensing convergence along lines of sight
through the Millenium Simulation to test this approach, and to calibrate
estimates of the total convergence along a line of sight to an observed
distant galaxy made by summing the convergences due to each object in a
photometric catalogue. Having found that the reconstruction process is effective given perfect
knowledge of halo mass and redshift, we investigated the effects of reasonable
uncertainties in the stellar mass and redshift of each halo, and propogated
these uncertainties into a $\pr\left(\sumkappah\right)$ for each line of
sight. We draw the following conclusions:

\begin{itemize}

\item Our model uses a truncated spherical NFW profile for
each dark matter halo and neglects voids, but despite the model's simplicity
the reconstructed $\sumkappah$ is a good tracer of $\kappax$.  We found that
with perfect knowledge of the halo mass and redshift (from the Millenium
Simulation catalogs), the reconstruction gives an unbiased estimate of
$\pr(\kappax)$ for a single line of sight that is almost almost a factor of 2
less broad than the global $\pr(\kappax)$.

\item  For the most overdense lines of sight, the reconstruction produces a
very broad PDF, but since the reconstruction can be performed before follow-up
time is invested it will be possible to discard the most uncertain lines of
sight and prevent the waste of follow-up time.

\item With complete spectroscopic redshift coverage and just an empirical
stellar mass to halo mass relation, we find that the median of
$\pr\left(\sumkappah\right)$ is a useful indicator for generating an estimate
of $\pr(\kappax)$ from the ensemble of simulated lines of sight. The resulting
PDF tends to be around 10\% broader than it would have been given perfect
knowledge of both halo mass and redshift; given only photometric redshifts
(which in turn give rise to much less certain stellar masss estimates) causes
another $\sim$10\% increase to the width of $\pr(\kappax)$.

\item It is very rare for halos further than 2 arcminutes to make a
significant contribution to $\kappax$. We also found that including halos
whose host galaxy is less luminous than $i=24$ does not significantly improve
our reconstruction proceedure. A photometric survey to this depth of a
4$\times$4 arcminute patch around the lens would approach the limiting
uncertainties of our simple reconstruction recipe, and yield a 
$\pr(\kappax)$ that has, on average, a width of 0.0163 --
50\% less broad than the global $\pr(\kappax)$.

\item We find that the lines of sight with the sharpest $\pr(\kappax)$ are
typically under-dense. With a photometric reconstruction of all lens fields,
and following up only the lenses with the most constraining $\pr(\kappax)$,
the width of the dominant statistical uncertainty in time-delay cosmography
can be halved, whilst at the same time decreasing any potentail for a
systematic error due to lenses being biased in $\kappax$.

\end{itemize}

\todo{Tom}{Add more conclusions about bias in a large sample of lenses...}

\todo{Tom}{Wrap up}

@@

\todo{Tom}{More on shear here.}
